---
title: 'Equivalence testing for SIP datasets'
output:
  html_notebook: default
  html_document:
  pdf_document: default
---

The goal is to select genes that are equivalent or not-expressed in one mating type and significant DE (with  the same cutoff value) in the other mating type.

The model formula of the SIP RNA-seq experiment with 3 timepoints is: 
$$ log(\hat{Y}f_{i}) = \beta_{c0} + \beta_{c15}X_{c15} + \beta_{t15}X_{t15} + \beta_{c1}X_{c1} + \beta_{t1}X_{t1} + \beta_{c3}X_{c3} + \beta_{t3}X_{t3} $$

# Equivalence analysis in MT+

### Load data

```{r include = FALSE}
library(edgeR) ; library(stageR) ; library(tximport) ; library(readr); library(ggplot2);library(tidyr) ; library(openxlsx)

wd <- getwd()
dir <- dirname(dirname(dirname(wd)))

quantFiles = list.files(paste0(dir,"/Mapping_output_Salmon/Mapping_MT+/"))

load(paste0(dir,"/Metadata/tx2gene.rda"))
data = tximport(files=paste0(dir,"/Mapping_output_Salmon/Mapping_MT+/",quantFiles,"/quant.sf"), type="salmon", txOut=FALSE, tx2gene=tx2gene, importer=read_tsv)
countsGust=data$counts
colnames(countsGust)=quantFiles
colnames(countsGust)[29] <- "3c0a" #was misnamed 

## remove technical replicate from run 2
countsGust <- countsGust[,!colnames(countsGust)%in%"1c0a2"]
 # Load data
d <- DGEList(countsGust)
 # filter on cpm
keep <- rowSums(cpm(d)>1)>=3
dont_keep <- rowSums(cpm(d)>1)< 3
low_expression <- names(dont_keep)[dont_keep]
d <- d[keep,]
table(keep) # number of genes after filtering on cpm 
 # normalization 
d$samples$lib.size <- colSums(d$counts)
d <- calcNormFactors(d)
 ### Design:
week=substr(colnames(d),1,1)
treat=substr(colnames(d),2,2)
time=substr(colnames(d),3,nchar(colnames(d))-1)
treatTime=factor(paste0(treat,"_",time))
run=factor(
    c(1, #1c0a
      1, #1c15a
      1, #1c1b
      1, #1c3b
      1, #1c6b
      1, #1c9b
      1, #1t15a
      1, #1t1a
      1, #1t3b
      1, #1t6b
      1, #1t9b
      2, #2c0b
      2, #2c15a
      1, #2c15b
      2, #2c1a
      2, #2c3b
      2, #2c6a
      2, #2c9b
      2, #2t15a
      2, #2t1a
      2, #2t3a
      2, #2t6b
      2, #2t9a
      2, #3c1b
      1, #3c3a
      2, #3c6c
      1, #3c9c
      2, #3c0a
      1, #3t15c
      2, #3t1c
      1, #3t3a
      2, #3t6b
      1)) #3t9a

 design <- model.matrix(~treatTime+week+run,contrasts.arg=list(week=contr.sum, run=contr.sum))
rownames(design) <- colnames(d)
 # Fitting the 'standard' model to extract the coefficients later
d <- estimateDisp(d,design)
plotBCV(d)
fit <- glmFit(d,design)
 ### Contrast matrix for only t15,t1 and t3
L=matrix(0,nrow=ncol(coef(fit)),ncol=3)
rownames(L)=colnames(coef(fit))
colnames(L)=c("t15","t1","t3")
L[c(7,2)+1,1] <- c(1,-1) #t15
L[c(6,1)+1,2] <- c(1,-1) #t1
L[c(8,3)+1,3] <- c(1,-1) #t3
```


## Equivalence testing with glmFit (LR)

```{r}
# function that returns a data frame of equivalence p-values (not controlled) for the three contrasts at a given FC cutoff
equivalence_test3 <- function(deltalog){
  for (contrast_number in c(1,2,3)){
    new_design <- contrastAsCoef(design, contrast = L[,contrast_number], first = TRUE)
    new_design <- data.frame(new_design$design)
    
    y <- d
    
    # get original offsets (existing offset is used to correct for library size and RNA composition effect)
    orig.offs <- getOffset(y)
    # add log(cutoff value for FC) to the offset of the samples we want to compare: this will make that we don't test for Ha =!= 0, but Ha =!= FC_value (two sided). 
    y$offset <- orig.offs + (deltalog)*new_design$C1
    # refit model with adjusted offset
    fit.up <- glmFit(y, new_design)
    # LRT
    lrt.up <- glmLRT(fit.up, coef = 1)
    # other side
    y$offset <- orig.offs - (deltalog)*new_design$C1
    # refit model with adjusted offset
    fit.down <- glmFit(y, new_design)
    # LRT
    lrt.down <- glmLRT(fit.down, coef = 1)
    # tests were two-sided. convert to one-sided.
    p.up <- ifelse(lrt.up$table$logFC < 0, lrt.up$table$PValue/2, 1)
    p.down <- ifelse(lrt.down$table$logFC > 0, lrt.down$table$PValue/2, 1)
    # get TOST p-value.
    p.tost2 <- pmax(p.up, p.down)
    
    assign(paste0("genes_", contrast_number), p.tost2)
  }
  genes123 <- data.frame(rownames(d),genes_1, genes_2, genes_3)
  colnames(genes123) <- c("gene_ID", "p_15", "p_1", "p_3")
  
  # first apply FWER (Holm) to the contrasts
  eq3_FWER <- data.frame(stringsAsFactors = F)
  for (row in 1:nrow(genes123)){                     # this should be possible with apply but I couldn't let it work :(
    eq3_FWER <- rbind(eq3_FWER, p.adjust(genes123[row,2:4], method = "holm"))
  }
  
  rownames(eq3_FWER) <- genes123$gene_ID
  colnames(eq3_FWER) <- colnames(eq3_FWER)[2:4]
  
  # select least significant contrast (t15,t1,t3)
  max_eq3_FWER <- apply(eq3_FWER,1,max)
  
  # FDR correction on least significant contrast
  max_eq3_FWER_FDR <- p.adjust(max_eq3_FWER, "fdr")
  
  sum(max_eq3_FWER_FDR <= 0.05)   # number of significant genes in all three timepoints 
 
  return(names(max_eq3_FWER_FDR[max_eq3_FWER_FDR <= 0.05]))
}

## Run the test
delta <- 3
deltalog <- log(delta)
deltalog2 <- log2(delta)

equiv3_LRT <- equivalence_test3(deltalog)
```

# Check results
```{r}
logFcMatrix <-  as.data.frame(read.delim(paste0(dir,"/DE_matrices/log2_Fold_Changes_MT+.txt"), header = TRUE, sep = "", row.names = 1, stringsAsFactors = FALSE))

log2FC <- logFcMatrix[rownames(logFcMatrix) %in% equiv3_LRT,]

log2FC <- data.frame( c(log2FC$t15, log2FC$t1, log2FC$t3), c(rep("T15", nrow(log2FC)), rep("T1", nrow(log2FC)), rep("T3", nrow(log2FC))))
colnames(log2FC) <- c("log2FC", "tp")
log2FC$tp <- factor(log2FC$tp, levels = unique(log2FC$tp))

ggplot(data = log2FC) +
    geom_jitter(aes(x = 1, y = log2FC), size = 1, alpha = 0.30) +
    geom_hline(aes(yintercept = 0), size= 1, linetype = "dashed", color = "red") +
  geom_hline(aes(yintercept = deltalog2), size = 0.5, color = "red")+
  geom_hline(aes(yintercept = -deltalog2), size = 0.5, color = "red") +
   xlab("") +
   ylab("Log2(FC)") + 
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
 facet_grid(~ tp, scales='free', space = "free_x") +
  theme_bw() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  scale_x_continuous(breaks = c(1)) +
  scale_y_continuous(breaks = c(-1,0,1))

ggplot(data = log2FC) +
    geom_violin(aes(x = 1, y = log2FC), size = 1, fill = "cyan4") +
    geom_hline(aes(yintercept = 0), size= 1, linetype = "dashed", color = "white") +
  geom_hline(aes(yintercept = deltalog2), size = 0.5, color = "red")+
  geom_hline(aes(yintercept = -deltalog2), size = 0.5, color = "red") +
   xlab("") +
   ylab("Log2(FC)") + 
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
 facet_grid(~ tp, scales='free', space = "free_x") +
  theme_bw()+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  scale_x_continuous(breaks = c(1)) +
  scale_y_continuous(breaks = c(-1,0,1))
```

###  Exporting "non-responsive" genes to take to the other mating type

Non-responsive genes are: (1) "low expression genes" (filtered) (2) genes equivalent at all three tp for FC cutoff x
```{r}
non_responsive <- data.frame(c(low_expression, equiv3_LRT), c(rep("filtered", length(low_expression)), rep("equivalent", length(equiv3_LRT))))
colnames(non_responsive) <- c("gene_ID","type")
```

# Analysis of DE in MT-

Now we switch to MT- and look which of the genes in 'non-responsive' are significantly DE (>< FC cutoff). 

### loading libraries and data 
```{r include = FALSE }
wd <- getwd()
dir <- dirname(dirname(dirname(wd)))

quantFilesSara = list.files(paste0(dir,"/Mapping_output_Salmon/Mapping_Moeys/"))
load(paste0(dir,"/Metadata/tx2gene.rda"))

## try http:// if https:// URLs are not supported
dataSara = tximport(files=paste0(dir,"/Mapping_output_Salmon/Mapping_Moeys/",quantFilesSara,"/quant.sf"), type="salmon", txOut=FALSE, tx2gene=tx2gene, importer=read_tsv)
countsSara=dataSara$counts
colnames(countsSara) = quantFilesSara
## sum counts across lanes for Sara's dataset.
countsSara = countsSara[,seq(1,41,by=2)] + countsSara[,seq(2,42,by=2)]

conditions <- read.delim(paste0(dir, "/Metadata/conditions_Moeys.txt"))
rownames(conditions) <- unlist(lapply(strsplit(as.character(conditions$files),split="_"),function(x) paste(x[1:2],collapse = "_")))
timeSara = conditions$time[match(colnames(countsSara),rownames(conditions))]
timeSara = as.character(timeSara)
timeSara[timeSara=="3h"] <- 3
timeSara[timeSara=="1h"] <- 1
timeSara[timeSara=="15m"] <- 0.25
timeSara <- as.numeric(timeSara)
treatSara = conditions$treatment[match(colnames(countsSara),rownames(conditions))]
treatSara <- (c("t","c"))[as.numeric(treatSara)] #control-treatment coding as in Gust's experiment

metaData = xlsx::read.xlsx(paste0(dir, "/Metadata/metadata_Moeys.xlsx"), sheetIndex=1, rowIndex=1:22, colIndex=1:9, header=TRUE)
metaData$UNIQUE.SAMPLE.NUMBER = as.character(metaData$UNIQUE.SAMPLE.NUMBER)
metaData$UNIQUE.SAMPLE.NUMBER = sub(x=metaData$UNIQUE.SAMPLE.NUMBER ,pattern="[a-z]+",replacement="")
metaData$UNIQUE.SAMPLE.NUMBER = as.numeric(metaData$UNIQUE.SAMPLE.NUMBER)
conditions$lib=as.factor(paste("L",sapply(rownames(conditions),function(x) substr(strsplit(x,"_")[[1]][1],5,7)),sep=""))
target1 = conditions[conditions$lane==1,]
target1$lib = sub(x=target1$lib,pattern="L",replacement="")
allData = merge(x=metaData, y=target1, by.x="UNIQUE.SAMPLE.NUMBER", by.y="lib")

conditions$lib2 =  sub(x=conditions$lib,pattern="L",replacement="")
matchID = match(x=conditions$lib2,table=allData$UNIQUE.SAMPLE.NUMBER)
conditions$week = allData$SAMPLE.NAME[matchID]
conditions$week = as.character(conditions$week)
conditions$week = factor(as.numeric(sapply(1:length(conditions$week), function(i) substr(conditions$week[i],start=nchar(conditions$week[i]),stop=nchar(conditions$week[i])))))
weekSara <- as.character(conditions$week[match(colnames(countsSara),rownames(conditions))])

###### design matrix
treatTime <- factor(paste0("t",timeSara,"_",treatSara))
treatTime = relevel(treatTime,ref="t0_c")
design <- model.matrix(~treatTime+weekSara)
```

### filtering (1) non-responsive genes MT+, then (2) cpm >1

```{r}
d <- DGEList(countsSara)

### COMMENTED OUT FOR INTERSECTION.
#(1) filter non-responsive genes
# d <- d[rownames(d$counts) %in% non_responsive$gene_ID,]
# nrow(d$counts) # number of genes after filtering of MT+ non-responsive genes

#(2) filter on cpm
keep <- rowSums(cpm(d)>1)>=3
d <- d[keep,]
nrow(d$counts) # number of genes after filtering on cpm 

# normalization 
d$samples$lib.size <- colSums(d$counts)
d <- calcNormFactors(d)

# Fit normal model without reparametrization
d <- estimateDisp(d,design)
plotBCV(d)

# fit model (based on the design factors and negative binomial dispersions, this will fit the negative binomial model through every level)
fit <- glmFit(d,design)

## check if the residuals follow a negative binomial distribution (Q-Q plot)
gofStat <- gof(fit,plot=TRUE)

## contrast matrix  (contrasts are used to determine which factors from the design matrix to compare, i.e. is the difference in coefficient significantly different from zero instead of 'is the coefficient different from zero', in this case compare treat vs control for every timepoint)

L <- matrix(0,nrow=ncol(fit$coefficients),ncol=3)
dimnames(L) <- list(colnames(fit$coefficients),c("t15","t1","t3"))
L[c("treatTimet0.25_t","treatTimet0.25_c"),1] <- c(1,-1)
L[c("treatTimet1_t","treatTimet1_c"),2] <- c(1,-1)
L[c("treatTimet3_t","treatTimet3_c"),3] <- c(1,-1)

```

### GLMtreat in MT-
Stagewise testing with P-value aggregation with the Sidak method

```{r}
library(aggregation)

tr15 <- glmTreat(fit, contrast = as.data.frame(L)$t15, lfc= deltalog2)
tr1 <- glmTreat(fit, contrast = as.data.frame(L)$t1, lfc= deltalog2)
tr3 <- glmTreat(fit, contrast = as.data.frame(L)$t3, lfc= deltalog2)

# Aggregate p-values
all_p_tr <- data.frame(tr15$table$PValue, tr1$table$PValue, tr3$table$PValue)
colnames(all_p_tr) <- c("t15", "t1", "t3")
rownames(all_p_tr) <- rownames(tr15)

p_screening_tr <- apply(all_p_tr, 1, function(x) sidak(x))
names(p_screening_tr) <- rownames(tr15)
p_confirmation_tr <- as.matrix(all_p_tr)

stageRobj_tr <- stageR(pConfirmation = p_confirmation_tr, pScreen = p_screening_tr)
stageRobj2_tr <- stageWiseAdjustment(object = stageRobj_tr, method = "holm", alpha = 0.05)

res_tr <- as.data.frame(getResults(stageRobj2_tr))

colSums(res_tr)

sign_ML <- res_tr[res_tr$padjScreen == 1,]

names <- rownames(res_tr[res_tr$padjScreen == 1,])

table(non_responsive$type[non_responsive$gene_ID %in% names])

log2FcMatrix_min <-  as.data.frame(read.delim(paste0(dir,"/DE_matrices/log2_Fold_Changes_MT-_Moeys.txt"), header = TRUE, sep = "", row.names = 1, stringsAsFactors = FALSE))


logFC_ML <- log2FcMatrix_min[rownames(log2FcMatrix_min) %in% names,]

res <- res_tr[res_tr$padjScreen == 1, 2:4]
logFC <- logFC_ML[,1:3] 

dir <- logFC*res

upDown <- function(value){
if (value > 0) {
  return("up")
} 
if (value == 0) {
  return("")
} 
else {
  return("down")
}
}

upD15 <- sapply(dir$t15, function(x) upDown(x))
upD1 <- sapply(dir$t1, function(x) upDown(x))
upD3 <- sapply(dir$t3, function(x) upDown(x))

out <- data.frame(names, upD15, upD1, upD3)

# Get the up/down response over all three timepoints 
out$upDall <- paste0(out$upD15, out$upD1, out$upD3 )

out$upDall[out$upDall== "upup"] <- "up"
out$upDall[out$upDall== "upupup"] <- "up"

out$upDall[out$upDall== "downdown"] <- "down"
out$upDall[out$upDall== "downdowndown"] <- "down"

out
```

## Intersection between MT+ and MT- results

```{r}
minGenes <- as.character(out$names)
plusGenes <- as.character(non_responsive$gene_ID)
interGenes <- intersect(minGenes, plusGenes)
write.table(interGenes, file="intersectionSRMGenes.txt", quote=FALSE, row.names = FALSE, col.names = FALSE)
```





### Plotting results for validation

```{r}

## MT+
logFC_ML <- logFcMatrix[rownames(logFcMatrix) %in% rownames(sign_ML),]
logFC_ML2 <- gather(logFC_ML, timepoint, log2FC, 1:3)
logFC_ML2$gene_ID <- rep(rownames(logFC_ML),3)
logFC_ML2$MT <- "MT+"
logFC_ML2$significant <- 1
logFC_ML2 <- logFC_ML2[9:13]


## MT-
logFC_ML_min <- log2FcMatrix_min[rownames(log2FcMatrix_min) %in% rownames(sign_ML),]
logFC_ML2_min <- gather(logFC_ML_min, timepoint, log2FC, 1:3)
logFC_ML2_min$gene_ID <- rep(rownames(logFC_ML_min),3)
logFC_ML2_min$MT <- "MT-"
logFC_ML2_min$significant <- c(sign_ML$t15, sign_ML$t1, sign_ML$t3)


## Both
both_ML <- rbind(logFC_ML2, logFC_ML2_min)
both_ML <- both_ML[both_ML$significant == 1,]

logFC_ML2_min
both_ML$MT <- relevel(as.factor(both_ML$MT), "MT+")
both_ML$timepoint <- relevel(as.factor(both_ML$timepoint), "t15")

both_ML$timepoint <- as.character(both_ML$timepoint)
both_ML$timepoint[both_ML$timepoint == "t15"] <- "15min"
both_ML$timepoint[both_ML$timepoint == "t1"] <- "1h"
both_ML$timepoint[both_ML$timepoint == "t3"] <- "3h"
both_ML$timepoint <- factor(both_ML$timepoint, levels = unique(both_ML$timepoint))

ggplot() +
     geom_jitter(aes(y = (both_ML$log2FC), x = both_ML$timepoint, fill = both_ML$MT), shape = 21, size = 1.2,  width = 0.2, height = 0, stroke = 0.6) +
     geom_hline(yintercept=(deltalog2)) +
     geom_hline(yintercept=-(deltalog2)) +
     xlab("Timepoint") +
     ylab("log2(FC)") +
   labs(fill = "Mating Type") +
  theme_bw() +
#  ggtitle(paste0("Equivalent in MT+ (Wald), DE in MT- (glmTreat): ", nrow(sign_ML), " genes")) +
   theme(plot.title = element_text(size= 10))
```
