---
title: "SRB with FDR control"
author: "Koen Van den Berge"
date: "10/17/2019"
output: html_document
---

This script will look for genes that are DE from the delta FC cutoff in both MT+ and MT- and are this at the heart of the SIP response

# MT- (Sara Moeys)

## Import & change data
```{r include=FALSE}
library(edgeR) ; library(stageR) ; library(tximport) ; library(readr); library(xlsx) ; library(ggplot2) ; library(tidyr)

wd <- getwd()
dir <- dirname(dirname(dirname(wd)))

quantFilesSara = list.files(paste0(dir,"/Mapping_output_Salmon/Mapping_Moeys/"))
load(paste0(dir,"/Metadata/tx2gene.rda"))

## try http:// if https:// URLs are not supported
dataSara = tximport(files=paste0(dir,"/Mapping_output_Salmon/Mapping_Moeys/",quantFilesSara,"/quant.sf"), type="salmon", txOut=FALSE, tx2gene=tx2gene, importer=read_tsv)
countsSara=dataSara$counts
colnames(countsSara) = quantFilesSara
## sum counts across lanes for Sara's dataset.
countsSara = countsSara[,seq(1,41,by=2)] + countsSara[,seq(2,42,by=2)]

conditions <- read.delim(paste0(dir, "/Metadata/conditions_Moeys.txt"))
rownames(conditions) <- unlist(lapply(strsplit(as.character(conditions$files),split="_"),function(x) paste(x[1:2],collapse = "_")))
timeSara = conditions$time[match(colnames(countsSara),rownames(conditions))]
timeSara = as.character(timeSara)
timeSara[timeSara=="3h"] <- 3
timeSara[timeSara=="1h"] <- 1
timeSara[timeSara=="15m"] <- 0.25
timeSara <- as.numeric(timeSara)
treatSara = conditions$treatment[match(colnames(countsSara),rownames(conditions))]
treatSara <- (c("t","c"))[as.numeric(treatSara)] #control-treatment coding as in Gust's experiment

metaData = read.xlsx(paste0(dir, "/Metadata/metadata_Moeys.xlsx"), sheetIndex=1, rowIndex=1:22, colIndex=1:9, header=TRUE)
metaData$UNIQUE.SAMPLE.NUMBER = as.character(metaData$UNIQUE.SAMPLE.NUMBER)
metaData$UNIQUE.SAMPLE.NUMBER = sub(x=metaData$UNIQUE.SAMPLE.NUMBER ,pattern="[a-z]+",replacement="")
metaData$UNIQUE.SAMPLE.NUMBER = as.numeric(metaData$UNIQUE.SAMPLE.NUMBER)
conditions$lib=as.factor(paste("L",sapply(rownames(conditions),function(x) substr(strsplit(x,"_")[[1]][1],5,7)),sep=""))
target1 = conditions[conditions$lane==1,]
target1$lib = sub(x=target1$lib,pattern="L",replacement="")
allData = merge(x=metaData, y=target1, by.x="UNIQUE.SAMPLE.NUMBER", by.y="lib")

conditions$lib2 =  sub(x=conditions$lib,pattern="L",replacement="")
matchID = match(x=conditions$lib2,table=allData$UNIQUE.SAMPLE.NUMBER)
conditions$week = allData$SAMPLE.NAME[matchID]
conditions$week = as.character(conditions$week)
conditions$week = factor(as.numeric(sapply(1:length(conditions$week), function(i) substr(conditions$week[i],start=nchar(conditions$week[i]),stop=nchar(conditions$week[i])))))
weekSara <- as.character(conditions$week[match(colnames(countsSara),rownames(conditions))])
```

## filter and normalize

```{r}
keep <- rowSums(cpm(countsSara)>1)>=3
table(keep)
countsSara <- countsSara[keep,]

d <- DGEList(countsSara)
d <- calcNormFactors(d)
```

## model design

```{r}
treatTime <- factor(paste0("t",timeSara,"_",treatSara))
treatTime = relevel(treatTime,ref="t0_c")
design <- model.matrix(~treatTime+weekSara)

colnames(countsSara) <- treatTime
write.table(countsSara, "counts_Moeys_SIPplus.txt")
write.table(treatSara, "treatment_Moeys_SIPplus.txt", row.names = F, col.names = F, quote =F)
write.table(timeSara, "time_Moeys_SIPplus.txt",  row.names = F, col.names = F, quote =F)
write.table(weekSara, "batch_effect_Moeys_SIPplus.txt",  row.names = F, col.names = F, quote =F)
```

## MDS plot

```{r}
t<- plotMDS(d,col=as.numeric(factor(timeSara)),pch=c(9,20)[as.numeric(factor(treatSara))])
legend("center", c("0m","15m","1h","3h"), pch=20, col=as.numeric(factor(timeSara))[1:6])

treat2 <- treatSara
treat2[treat2 == "c"] <- "control"
treat2[treat2 == "t"] <- "SIP+ treated"

time2 <- as.factor(timeSara)


MDS_df <- data.frame(t$x, t$y)

ggplot(data = MDS_df) +
  geom_point(aes(x = t.x, y = t.y,  fill = time2,  shape = treat2), size = 3)  +
  theme_classic() +
  theme(legend.box="horizontal") +
  scale_shape_manual(values = c(21,23))+
  scale_fill_manual(values= c("grey20",  "orange","firebrick3", "olivedrab3")) +
  guides(fill = guide_legend(override.aes=list(shape=21)), shape = guide_legend(override.aes=list(fill = "black"))) +
  theme(legend.position = c(0.5, 0.55)) +
  labs(fill = "Time (h)", shape = "Treatment") +
  xlab("Leading logFC dim1") +
  ylab("Leading logFC dim2")

##ggsave("plots/MDS_min.pdf" ,width = 6, height = 4)

```

## model fitting

```{r}
d <- DGEList(countsSara)
d <- calcNormFactors(d)
d <- estimateDisp(d,design)
plotBCV(d)

fit <- glmFit(d,design)

## check if the residuals follow a negative binomial distribution (Q-Q plot)
gofStat <- gof(fit,plot=TRUE)
```

## Stagewise testing with Sidak p-value aggregation (as glmTreat does not handle omnibus tests we use sidak p-value aggregation for omnibus testing)

```{r}
library(aggregation)

L <- matrix(0,nrow=ncol(fit$coefficients),ncol=3)
dimnames(L) <- list(colnames(fit$coefficients),c("t15","t1","t3"))
L[c("treatTimet0.25_t","treatTimet0.25_c"),1] <- c(1,-1)
L[c("treatTimet1_t","treatTimet1_c"),2] <- c(1,-1)
L[c("treatTimet3_t","treatTimet3_c"),3] <- c(1,-1)

tr15 <- glmTreat(fit, contrast = as.data.frame(L)$t15, lfc=log2(3))
tr1 <- glmTreat(fit, contrast = as.data.frame(L)$t1, lfc=log2(3))
tr3 <- glmTreat(fit, contrast = as.data.frame(L)$t3, lfc=log2(3))

# Aggregate p-values
all_p_tr <- data.frame(tr15$table$PValue, tr1$table$PValue, tr3$table$PValue)
colnames(all_p_tr) <- c("t15", "t1", "t3")
rownames(all_p_tr) <- rownames(tr15)

pScreenMin <- apply(all_p_tr, 1, function(x) sidak(x))
names(pScreenMin) <- rownames(tr15)
p_confirmation_tr <- as.matrix(all_p_tr)

stageRobj_tr_min <- stageR(pConfirmation = p_confirmation_tr, pScreen = pScreenMin)
stageRobj2_tr_min <- stageWiseAdjustment(object = stageRobj_tr_min, method = "holm", alpha = 0.05)

res_tr_min <- as.data.frame(getResults(stageRobj2_tr_min))

colSums(res_tr_min)

minGenes <- rownames(res_tr_min)[res_tr_min[,1] == 1]
```


# MT + 

# Import

```{r include=FALSE}
library(ggplot2)
library(tidyr)

wd <- getwd()
dir <- dirname(dirname(dirname(wd)))

quantFiles = list.files(paste0(dir,"/Mapping_output_Salmon/Mapping_MT+/"))

load(paste0(dir,"/Metadata/tx2gene.rda"))
data = tximport(files=paste0(dir,"/Mapping_output_Salmon/Mapping_MT+/",quantFiles,"/quant.sf"), type="salmon", txOut=FALSE, tx2gene=tx2gene, importer=read_tsv)
countsGust=data$counts
colnames(countsGust)=quantFiles
colnames(countsGust)[29] <- "3c0a" #was misnamed 

## remove technical replicate from run 2
countsGust <- countsGust[,!colnames(countsGust)%in%"1c0a2"]
 # Load data
d <- DGEList(countsGust)
 # filter on cpm
keep <- rowSums(cpm(d)>1)>=3
dont_keep <- rowSums(cpm(d)>1)< 3
low_expression <- names(dont_keep)[dont_keep]
d <- d[keep,]
table(keep) # number of genes after filtering on cpm 
 # normalization 
d$samples$lib.size <- colSums(d$counts)
d <- calcNormFactors(d)
 ### Design:
week=substr(colnames(d),1,1)
treat=substr(colnames(d),2,2)
time=substr(colnames(d),3,nchar(colnames(d))-1)
treatTime=factor(paste0(treat,"_",time))
run=factor(
    c(1, #1c0a
      1, #1c15a
      1, #1c1b
      1, #1c3b
      1, #1c6b
      1, #1c9b
      1, #1t15a
      1, #1t1a
      1, #1t3b
      1, #1t6b
      1, #1t9b
      2, #2c0b
      2, #2c15a
      1, #2c15b
      2, #2c1a
      2, #2c3b
      2, #2c6a
      2, #2c9b
      2, #2t15a
      2, #2t1a
      2, #2t3a
      2, #2t6b
      2, #2t9a
      2, #3c1b
      1, #3c3a
      2, #3c6c
      1, #3c9c
      2, #3c0a
      1, #3t15c
      2, #3t1c
      1, #3t3a
      2, #3t6b
      1)) #3t9a

design <- model.matrix(~treatTime+week+run,contrasts.arg=list(week=contr.sum, run=contr.sum))
rownames(design) <- colnames(d)

design
```

# Dispersion estimation, model fitting

```{r}
## estimate dispersion
d <- estimateDisp(d,design)
plotBCV(d)

## fit model
fit <- glmFit(d,design)
fitPlus <- fit

## check if the residuals follow a negative binomial distribution (Q-Q plot)
gofStat <- gof(fit,plot=TRUE)
```

# Stage-wise hypothesis testing using stageR

```{r}
## contrast matrix
L=matrix(0,nrow=ncol(coef(fit)),ncol=5)
rownames(L)=colnames(coef(fit))
colnames(L)=c("t15","t1","t3","t6","t9")#,
	      #"avFC","lFC1","lFC2","eFC",
	      #"lFC1-eFC","lFC2-eFC")
L[c(7,2)+1,1] <- c(1,-1) #t15
L[c(6,1)+1,2] <- c(1,-1) #t1
L[c(8,3)+1,3] <- c(1,-1) #t3
```


```{r}
# Test all genes
tr15 <- glmTreat(fitPlus, contrast = as.data.frame(L)$t15, lfc= log2(3))
tr1 <- glmTreat(fitPlus, contrast = as.data.frame(L)$t1, lfc= log2(3))
tr3 <- glmTreat(fitPlus, contrast = as.data.frame(L)$t3, lfc= log2(3))

# Aggregate p-values
allPPlus <- data.frame(tr15$table$PValue, tr1$table$PValue, tr3$table$PValue)
colnames(allPPlus) <- c("t15", "t1", "t3")
rownames(allPPlus) <- rownames(tr15)

pScreenPlus <- apply(allPPlus, 1, function(x) sidak(x))
names(pScreenPlus) <- rownames(tr15)
pConfirmationPlus <- as.matrix(allPPlus)

objPlus <- stageR(pConfirmation = pConfirmationPlus, pScreen = pScreenPlus)
objPlus <- stageWiseAdjustment(object = objPlus, method = "holm", alpha = 0.05)

resPlus <- as.data.frame(getResults(objPlus))
colSums(resPlus)
plusGenes <- rownames(resPlus)[resPlus[,1]==1]
```

# Intersection

```{r}
bothGenes <- intersect(plusGenes, minGenes)
write.table(bothGenes, file="intersectionBothGenes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```




